---
  title: "usage deseqAbstraction - iNd24 example"
author: "Per Ludvik Brattaas"
date: "November 7, 2017"
output:
  html_document: default
pdf_document:
  latex_engine: xelatex
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Tutorial on deseqAbstraction
## Analysing iN-d24 RNA-sequencing data

### Easy analysis and visualization of featureCount 

#### 1. Setup of workspace


```{r setup workspace, message = FALSE}

# clean environment and workspace
rm(list=ls())
gc()

# load deseqAbstraction and DESeq2 package
library(devtools)
install_github("perllb/deseqAbstraction")
library(deseqAbstraction)
library(DESeq2)

# path to project
path <- "/home/pbrattaas/LocalData/iN-Janelle/"

# path and filename of raw featureCounts file
fc.file <- "Data/Quant/hg38.mMap.NCBI.Exon.txt"
fc.file.path <- paste(path,fc.file,sep = "")
```


#### 2. Define sample names (Highly recommended to do this properly, this is a bad example with bad filenames)
```{r}

head <- read.delim(file = fc.file.path,header = T,nrows = 1,skip = 1)
# 1. take the column names of the rawCount table from featureCounts
cnames <- colnames(head[,-c(1:6)])

# 2. remove the substrings that are not needed for identifying the samples, noise from filenames
newnames <- gsub(pattern = "X.projects.fs1.medpvb.backup.projects.iN.d24.Aligned_hg38_STAR_mMap.hg38.mMap.",replacement = "",x = cnames)
newnames <- gsub(pattern = "_L004Aligned.sortedByCoord.out.bam",replacement = "",x = newnames)
newnames <- gsub(pattern = "_L005Aligned.sortedByCoord.out.bam",replacement = "",x = newnames)
sampleNames <- gsub(pattern = "_L006Aligned.sortedByCoord.out.bam",replacement = "",x = newnames)

```


#### 3. colData data.frame that describes the data

```{r}
# make one vector describing if the sample has REST inhibition or not
rest <- ifelse((grepl("REST",sampleNames)),yes = "RESTi",no="")
# cell type of sample?
cell <- c(rep("C2",18),rep("HFL1",6))
# describe vector construct of sample
vector <- c(rep("pB_miR_pA",3),rep("pBpA",3),rep("",3),rep(c("pBpA","pB+pA"),3),rep("pB_miR_pA",3),rep("",3),rep("pA+pB",3))
# condition (main design vector) had all info
cond <- paste(cell,vector,rest,sep = "_")
# make a data.frame with these vectors and store as colData
# here, you always should have the main design as "condition" 
colData <- data.frame(condition=cond,cell=cell,vector=vector,rest=rest,samples=sampleNames)


```

### 4. create deseqAbs object
```{r }
# generate deseqAbs object
iN <- deseqAbs$new(name = "iN-d24",filename = fc.file.path,colData=colData)

```


#### 5.  Run DESeq pipeline
##### - You can try to do this in one function $fullAuto(). 
##### - Full auto has four main steps:
##### (1) - create DEseq object (access with $deseq)
##### (2) - Do diffex analysis (access with $test)
##### (3) - Do varianceStablizingTransformation (access with $VST)
##### (4) - Do RPKM normalization (access with $rpkm)

##### - However, for this automation, it is critical that your deseqAbs object has
##### (1) - the raw featureCount file in the correct format
##### (2) - the colData data.frame with at least one column called "condition", which contain the grouping of your samples, that will later be used for diffex anaysis

```{r}
## run pipeline
iN$fullAuto()

## by default, fullAuto do blind dispersion estimates, which is good if relatively few genes are expected to have very different counts in different conditions.
## if many genes are expected to have differential expression, it might be better to do non-blind estimates, which can be manually done by calling makeVST() with blind = F.
iN$makeVST(blind=T)

# take a peak at normalized data generated
# rpkm
head(iN$FPKM)
# varianceStablizingTransformed
head(iN$VST)
# default diffex
head(iN$test$Default)

```

#### 6. Visualize data
#### 6.1 PCA and sample-to-sample

```{r}

# plot PCA without labels on points
PCAplotter(dat = iN$VST,title = "NCBI top 5000",ntop = 5000,color = iN$colData$rest,shape=iN$colData$vector)

# plot PCA with labels
PCAplotter(dat = iN$VST,title = "NCBI top 5000",ntop = 5000,color = iN$colData$rest,shape=iN$colData$vector,label = iN$colData$cell)


```

#### 7. Differential expression
##### By default, the fullAuto() function makes a diffex analysis with the last level vs the first level of the last variables in the design formula colData (design vector)
##### However, if you have many conditions you may want to test other pairwise combinations

```{r}

## Create directory to print diffex result matrices
deseq.res.path <- paste(path,"Results/DEseq/",sep = "")
if(!dir.exists(deseq.res.path)) {
  
  dir.create(path = deseq.res.path)
  
}

```

#### 8. test pBpA+RESTi vs pBpA (no RESTi)
```{r }
# get the names of the conditions 
conds <- gsub(pattern = "condition",replacement = "",resultsNames(iN$deseq))
# list condition names with index
cbind(conds,1:length(conds))

# Test RESTi+pBpA (7th condition) vs. pBpA (6th condition) (without RESTi)
c1 <- conds[7]
c2 <- conds[6]

iN$makeDiffex(name = paste(c2,c1,sep=".vs."),c1 = c1,c2 = c2)
head(iN$test$RESTi)
# write diffex table sorted on p-adj
write.table(x = iN$test$RESTi[order(iN$test$RESTi$padj),],file = paste(deseq.res.path,c1,"_vs._",c2,".Diffex.txt",sep = ""),quote = F,sep = "\t")

meanPlot(ex = iN$baseMean$Mean[,c(c2,c1)],p = .00001,c1=c1,c2=c2,test = iN$test$C2_pBpA_RESTi.vs.C2_pBpA_)

maPlot(test = iN$test$C2_pBpA_RESTi.vs.C2_pBpA_,c1 = c1,c2 = c2,p = .0001)

```

#### 9. test pBpA+RESTi vs RESTi (no pBpA)
```{r }

# Test RESTi+pBpA  (7th) vs. RESTi (2nd ) (without pBpA)
c1 <- conds[7] #pBpA + RESTi
c2 <- conds[2] #RESTi only
iN$makeDiffex(name = "pBpA",c1 = c1,c2 = c2)
head(iN$test$pBpA)
# write diffex table sorted on p-adj
write.table(x = iN$test$pBpA[order(iN$test$pBpA$padj),],file = paste(deseq.res.path,c1,"_vs._",c2,".Diffex.txt",sep = ""),quote = F,sep = "\t")

meanPlot(exp = iN$baseMean$Mean[,c(c2,c1)],test = iN$test$pBpA,c1 = c1,c2 = c2,p = .0001)
maPlot(test = iN$test$pBpA,c1 = c1,c2 = c2,p = .0001)

```

#### 10. test pBpA+RESTi vs pBmirpA 
```{r }
cbind(conds,1:length(conds))

# Test RESTi+pBpA  (7th) vs. pBmirPA (3nd)
c1 <- conds[7] #pBpA + RESTi
c2 <- conds[3] #RESTi only
iN$makeDiffex(name = "pBmiRpA",c1 = c1,c2 = c2)
head(iN$test$pBmiRpA)

# write diffex table sorted on p-adj
write.table(x = iN$test$pBmiRpA[order(iN$test$pBmiRpA$padj),],file = paste(deseq.res.path,c1,"_vs._",c2,".Diffex.txt",sep = ""),quote = F,sep = "\t")

meanPlot(exp = iN$baseMean$Mean[,c(c2,c1)],test = iN$test$pBmiRpA,c1 = c1,c2 = c2,p = .0001)
maPlot(test = iN$test$pBmiRpA,c1 = c1,c2 = c2,p = .0001)


```


#### 11. test pBpA+RESTi vs pBmirpA+RESTi
```{r }
cbind(conds,1:length(conds))

# Test RESTi+pBpA  (7th) vs. pBmirPA (3nd)
c1 <- conds[7] #pBpA + RESTi
c2 <- conds[4] #RESTi only
iN$makeDiffex(name = "pBmiRpA.RESTi",c1 = c1,c2 = c2)
head(iN$test$pBmiRpA.RESTi)

# write diffex table sorted on p-adj
write.table(x = iN$test$pBmiRpA.RESTi[order(iN$test$pBmiRpA.RESTi$padj),],file = paste(deseq.res.path,c1,"_vs._",c2,".Diffex.txt",sep = ""),quote = F,sep = "\t")

meanPlot(exp = iN$baseMean$Mean[,c(c2,c1)],test = iN$test$pBmiRpA.RESTi,c1 = c1,c2 = c2,p = .0001)
maPlot(test = iN$test$pBmiRpA.RESTi,c1 = c1,c2 = c2,p = .00001)


```


#### 12. volcano plot

```{r}

# generate volcanoplot RESTi 
volcanoPlot(test = iN$test$RESTi)

# cut y-axis 
volcanoPlot(test = iN$test$RESTi,max = 100)

# generate volcanoplot pBpA
volcanoPlot(test = iN$test$pBpA,p = .0001)

# cut y-axis 
volcanoPlot(test = iN$test$pBpA,max = 100,p = .0001)


```

#### 13. heatmaps 1: Plot expression of most significantly changed genes (selected on p-adj from user-defined diffex-test) 
```{r}

# plot top significant genes in RESTi
mostSignificantHeat(data = assay(iN$VST),test = iN$test$RESTi)

# plot top significant genes, top 20
mostSignificantHeat(data = assay(iN$VST),test = iN$test$pBpA,ntop = 20)

# plot top significant genes, top 20, with annotation of columns
mostSignificantHeat(data = assay(iN$VST),test = iN$test$RESTi,ntop = 20,a1 = iN$colData$condition,n1 = "Condition")

## plot top significant genes, top 50, with two annotation of columns
mostSignificantHeat(data = assay(iN$VST),test = iN$test$RESTi,ntop = 50,a1 = iN$colData$vector,n1 = "Vector",a2 = iN$colData$rest,n2 = "Rest")

```

#### 5.4 heatmaps 2: Plot the most variable genes (over all samples given..)
```{r }

# plot most variable genes
mostVariableHeat(data = assay(iN$VST))

# plot most variable genes
mostVariableHeat(data = assay(iN$VST),ntop = 10,a1 = iN$colData$vector,n1 = "Vector",a2 = iN$colData$rest,n2 = "REST")


```

#### 5.4 heatmaps 3: plot a given set of genes

```{r}

genes <- c("LPPR4", "NEFH", "BOC", "TRPC6", "GPRIN1", "ST8SIA2", "PTK2B", "UCHL1", "CTNNA1", "PLCG1", "SEMA6C", "DSCAML1")

# sd defines cutoff for standard deviation (sd = 1 plots only genes with sd > 1). Set sd=-1 of all genes should be plotted
heatGenes(data = assay(iN$VST),genes = genes,sd = .5)

# if you want to normalize expression on each row (z-score), set z = T
heatGenes(data = assay(iN$VST),genes = genes,sd = -1,z = T)

# add column annotations
heatGenes(data = assay(iN$VST),genes = genes,sd = -1,z = F,a1 = iN$colData$vector,n1 = "Vector",a2 = iN$colData$rest,n2 = "REST")


```

#### Choose a subset of conditions/samples to plot

```{r}

## plot only C2 samples

# get index of those
idx <- grep("C2",iN$sampleNames)

# plot most variable genes - of the C2 samples
mostVariableHeat(data = assay(iN$VST)[,idx],ntop = 10,a1 = iN$colData$vector[idx],n1 = "Vector",a2 = iN$colData$rest[idx],n2 = "REST")

## plot only pBpA samples

# get index of those
idx <- grep("pBpA",iN$sampleNames)

# plot most variable genes - of the pBpA samples
mostVariableHeat(data = assay(iN$VST)[,idx],ntop = 30,a1 = iN$colData$vector[idx],n1 = "Vector",a2 = iN$colData$rest[idx],n2 = "REST")


```

#### 6. get average (and standard deviation) of expression levels of all genes within each condition

```{r}
base <- getAverage(iN$deseq)
head(base$Mean)
head(base$SD)

```

#### 7. Get significant genes (of RESTi test)
```{r}

# Get diffex-data of genes with p-adj < 0.001 and log2FC more/less than 0.2
sign <- getSign(x = iN$test$RESTi,p = .001,l = .2)
up <- sign$up
down <- sign$down

# Get names of genes with p-adj < 0.01 and log2FC more/less than .1
signID <- getSignName(x = iN$test$RESTi,p = .01,l = .2)
upID <- signID$up
downID <- signID$down

```

